---
title: "Bone-marrow pilot scRNA-seq of SA infected mice"
author: "Anna Hakobyan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, 
                      fig.width = 10, fig.height = 7)
```


```{r libs-load, include = FALSE}
#wget --recursive --no-parent -e robots=off -R "index.html*"  https://biomedical-sequencing.at/projects/BSA_0571_PS_BM_b361cecc39de49a097ae369fe636db81/ 

library(Seurat)
library(SeuratData)
library(SeuratDisk)

library(dplyr)
library(patchwork)
library(here)
library(ggpubr)
library(kableExtra)
library(tidyverse)

outdir = here("output")
data_dir = here("data/BSA_0571_PS_BM_b361cecc39de49a097ae369fe636db81")

if (! file.exists(outdir)) {
    dir.create(outdir)
}

# source(here("scripts", "marker_genes.R"))

```


```{r reading-in, include = FALSE}
hto_index = read.table(file.path(data_dir, "COUNT",
                                 "MR_30_10X_SA_PBS_BM_22112021_transcriptome", 
                                 "HTO_demux.csv"), h = T, sep = "," )
rownames(hto_index) = paste0(hto_index$index, "-1")
hto_index$index = NULL
colnames(hto_index) = "sample"

bm.data = Read10X(file.path(data_dir, "COUNT",
                              "MR_30_10X_SA_PBS_BM_22112021_transcriptome", 
                              "filtered_feature_bc_matrix"))

bm = CreateSeuratObject(counts  = bm.data$`Gene Expression`, 
                        meta.data = hto_index,
                        project = "BM",
                  min.cells = 3)

# bm.split = SplitObject(bm, split.by = "sample")
```

### HTO analysis
```{r}
bm[["HTO"]] = CreateAssayObject(counts = bm.data$`Antibody Capture`)
bm <- NormalizeData(bm, assay = "HTO", normalization.method = "CLR")
```

```{r}
# If you have a very large dataset we suggest using k_function = 'clara'. This is a k-medoid
# clustering function for large applications You can also play with additional parameters (see
# documentation for HTODemux()) to adjust the threshold for classification Here we are using
# the default settings
bm <- HTODemux(bm, assay = "HTO", positive.quantile = 0.99)
```


```{r}
table(bm$HTO_classification.global)
```

```{r}
# Group cells based on the max HTO signal
Idents(bm) <- "HTO_maxID"
RidgePlot(bm, assay = "HTO", features = rownames(bm[["HTO"]])[1:4], ncol = 2)
```

```{r}
Idents(bm) = "HTO_classification.global"
# First, we will remove negative cells from the object
bm.subset <- subset(bm, idents = "Negative", invert = TRUE)

# Calculate a tSNE embedding of the HTO data
DefaultAssay(bm.subset) <- "HTO"
bm.subset <- ScaleData(bm.subset, features = rownames(bm.subset),
    verbose = FALSE)
bm.subset <- RunPCA(bm.subset, features = rownames(bm.subset), approx = FALSE)
bm.subset <- RunTSNE(bm.subset, dims = 1:8, perplexity = 100, check_duplicates = FALSE)
DimPlot(bm.subset)
```

```{r}
# Extract the singlets
bm.singlet <- subset(bm, idents = "Singlet")

# Select the top 1000 most variable features
bm.singlet <- FindVariableFeatures(bm.singlet)

# Scaling RNA data, we only scale the variable features here for efficiency
bm.singlet <- ScaleData(bm.singlet, features = VariableFeatures(bm.singlet))

# Run PCA
bm.singlet <- RunPCA(bm.singlet, features = VariableFeatures(bm.singlet))
# We select the top 10 PCs for clustering and tSNE based on PCElbowPlot
bm.singlet <- FindNeighbors(bm.singlet, reduction = "pca", dims = 1:10)
bm.singlet <- FindClusters(bm.singlet, resolution = 0.6, verbose = FALSE)
bm.singlet <- RunTSNE(bm.singlet, reduction = "pca", dims = 1:10)

# Projecting singlet identities on TSNE visualization
DimPlot(bm.singlet, group.by = "HTO_classification")
```

```{r}
bm = bm.singlet
```



```{r, fig.width = 8}
total_counts_scatterplot = bm@meta.data %>% 
    arrange(nCount_RNA) %>% 
    group_by(sample) %>% 
    mutate(counter = row_number(sample)) %>% 
    ggplot(aes(x = counter, y = nCount_RNA, color = sample)) + geom_point()

ggsave(total_counts_scatterplot, filename = here("figures/total_nCount_RNA_scatter.png"),
       width = 7, height = 5)

violins_for_samples = bm@meta.data %>% 
    arrange(nCount_RNA) %>% 
    group_by(sample) %>% 
    ggplot(aes(x = sample, y = nCount_RNA, fill = sample)) + 
    geom_violin() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(violins_for_samples)
ggsave(violins_for_samples, filename = here("figures/total_nCount_RNA_violin.png"),
       width = 7, height = 5)
```


```{r mt-percent, include = FALSE}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
# for(smp in names(bm.split)) {
#     bm.split[[smp]][["percent.mt"]] = PercentageFeatureSet(bm.split[[smp]], pattern = "^mt-")
# }
bm[["percent.mt"]] = PercentageFeatureSet(bm, pattern = "^mt-")
```


### QC and normalization

The QC plots for number of features and number of counts, as well as percentage 
of mitochondrial features.

```{r qc-plots, echo = FALSE, warning = FALSE}
bm_qc = VlnPlot(bm, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                group.by = "sample", ncol = 3, log = TRUE) 
print(bm_qc)
```

Before filtering there are `r Cells(bm) %>% length` cells and `r nrow(bm)` genes.

```{r filt_params}
nfeatRNA_thresh_low = 150
nfeatRNA_thresh_up = 20000
ncount_min = 50
percentmt = 100
```

Filtering criteria are as follows:

1. min number of features(genes) - `r nfeatRNA_thresh_low`
2. max number of features(genes) - `r nfeatRNA_thresh_up`
3. min number of reads in a cell - `r ncount_min`
4. percent of mitochondrial genes - `r percentmt`


```{r filtering-out, include = FALSE}
init_bm_length = length(bm$nFeature_RNA)

bm <- subset(bm, subset = nFeature_RNA > nfeatRNA_thresh_low &
                 nFeature_RNA < nfeatRNA_thresh_up &
                 percent.mt < percentmt &
                 nCount_RNA > ncount_min
                 )

bm_qc = VlnPlot(bm, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                group.by = "sample", ncol = 3, log = TRUE) 

print(bm_qc)
# for(smp in names(bm.split)) {
#     cat(smp)
#     bm.split[[smp]] = subset(bm.split[[smp]], subset = nFeature_RNA > nfeatRNA_thresh_low &
#                  nFeature_RNA < nfeatRNA_thresh_up &
#                  percent.mt < percentmt &
#                  nCount_RNA > ncount_min)
# }

bm = subset(bm, subset = sample %in% c("HTO-PBS1", "HTO-PBS2", 
                                       "HTO-SA1", "HTO-SA2"))

bm$sample = gsub("^HTO-MR_30_10X_(.*)_BM.*$", "\\1", bm$sample)
```

After filtering there are `r Cells(bm) %>% length` cells and `r nrow(bm)` genes.

<!-- ```{r normalize-counts, include = FALSE, message=FALSE} -->
<!-- bm <- NormalizeData(bm, normalization.method = "LogNormalize", scale.factor = 10000) -->
<!-- #  -->
<!-- # for(smp in names(bm.split)) { -->
<!-- #     bm.split[[smp]] = NormalizeData(bm.split[[smp]], normalization.method = "LogNormalize", scale.factor = 10000) -->
<!-- # } -->
<!-- ``` -->

<!-- ### Identification of highly variable features -->

<!-- ```{r highly-variable-features-bm, echo = FALSE, warning=FALSE} -->
<!-- bm <- FindVariableFeatures(bm, selection.method = "vst", nfeatures = 2000) -->

<!-- # # Identify the 10 most highly variable genes -->
<!-- bm_top10 <- head(VariableFeatures(bm), 10) -->
<!-- bm_top30 <- head(VariableFeatures(bm), 30)   -->
<!-- bm_top30 %>% enframe(name = NULL, value = "top30Genes") %>% kbl() -->

<!-- # plot variable features with and without labels -->
<!-- bm_variable_genes_plot <- VariableFeaturePlot(bm) -->
<!-- bm_variable_genes_labeled <- LabelPoints(plot = bm_variable_genes_plot,  -->
<!--                                            points = bm_top30, repel = TRUE) +  -->
<!--     plot_annotation(title = "bm - highly variable genes") -->
<!-- print(bm_variable_genes_labeled) -->
<!-- ``` -->

<!-- <!-- ```{r highly-variable-features-bm-split, echo=FALSE, warning=FALSE, results='hide', fig.keep='all} --> -->
<!-- <!-- bm_variable_genes_plot.split = list() --> -->

<!-- <!-- for(smp in names(bm.split)) { --> -->
<!-- <!--     bm.split[[smp]] = FindVariableFeatures(bm.split[[smp]], selection.method = "vst",  --> -->
<!-- <!--                                            nfeatures = 2000) --> -->

<!-- <!--     vg_plot = VariableFeaturePlot(bm.split[[smp]]) --> -->
<!-- <!--     bm_variable_genes_plot.split[[smp]] =  --> -->
<!-- <!--         LabelPoints(plot = vg_plot, points = head(VariableFeatures(bm.split[[smp]]), 30), --> -->
<!-- <!--                     repel = TRUE) + plot_annotation(title = smp ) --> -->
<!-- <!-- } --> -->

<!-- <!-- highly_variable_regions_plots = ggarrange(plotlist = bm_variable_genes_plot.split, --> -->
<!-- <!--                                           nrow = 3, ncol = 2) --> -->
<!-- <!-- print(highly_variable_regions_plots) --> -->
<!-- <!-- ``` --> -->


<!-- Next, scaling is applied, which transforms gene expression to have a mean of  -->
<!-- 0 and a variance of 1 across cells. -->

<!-- ```{r scaling, include = FALSE} -->
<!-- bm_all_genes = rownames(bm) -->
<!-- bm = ScaleData(bm, features = bm_all_genes) -->

<!-- for(smp in names(bm.split)) { -->
<!--     bm.split[[smp]] = ScaleData(bm.split[[smp]], features = bm_all_genes) -->
<!-- } -->
<!-- ``` -->

### scTransforming

```{r}
bm = SCTransform(bm, method = "glmGamPoi", vars.to.regress = "percent.mt", verbose = FALSE)
```


```{r}
bm <- RunPCA(bm, verbose = FALSE)
bm <- RunUMAP(bm, dims = 1:30, verbose = FALSE)

bm <- FindNeighbors(bm, dims = 1:30, verbose = FALSE)
bm <- FindClusters(bm, verbose = FALSE)
DimPlot(bm, label = TRUE) + NoLegend()
```

```{r}
saveRDS(bm, file = file.path(outdir, "bm_pilot.rds") )
# bm = readRDS(file = file.path(outdir, "bm_pilot.rds"))
bm.split = SplitObject(bm, split.by = "sample")
saveRDS(bm.split, file = file.path(outdir, "bm_split_pilot.rds"))
```


<!-- ### Running linear dimensionality reduction(PCA) -->

<!-- ```{r pca, echo = FALSE, message=FALSE} -->
<!-- bm <- RunPCA(bm, features = VariableFeatures(object = bm)) -->

<!-- # for(smp in names(bm.split)) { -->
<!-- #     bm.split[[smp]] = RunPCA(bm.split[[smp]],  -->
<!-- #                              features = VariableFeatures(object = bm.split[[smp]])) -->
<!-- # } -->
<!-- ``` -->

<!-- ```{r determine-dimensionality, include = FALSE} -->
<!-- bm <- JackStraw(bm, num.replicate = 100, dims = 50) -->
<!-- bm <- ScoreJackStraw(bm, dims = 1:50) -->

<!-- JackStrawPlot(bm, dims = 1:50) # PC32  -->

<!-- # bm_3classes = subset(bm, subset = sample %in% c("HTO-MR_30_10X_SA1_BM_22112021", -->
<!-- #                                                 "HTO-MR_30_10X_PBS1_BM_22112021", -->
<!-- #                                                 "HTO-MR_30_10X_PBS2_BM_22112021")) -->
<!-- #  -->
<!-- # bm_3classes <- JackStraw(bm_3classes, num.replicate = 100, dims = 50) -->
<!-- # bm_3classes <- ScoreJackStraw(bm_3classes, dims = 1:40) -->
<!-- #  -->
<!-- # JackStrawPlot(bm_3classes, dims = 1:40) -->
<!-- ``` -->


<!-- ```{r finding-clusters, include = FALSE} -->
<!-- bm <- FindNeighbors(bm, dims = 1:32) -->
<!-- bm <- FindClusters(bm, resolution = 0.5) -->
<!-- ``` -->


<!-- ### Running non-linear dimensionality reduction with UMAP -->


<!-- ```{r, echo = FALSE} -->
<!-- bm <- RunUMAP(bm, dims = 1:32) -->

<!-- bm_umap = DimPlot(bm, reduction = "umap") +  -->
<!--     plot_annotation(title = "bm UMAP") -->

<!-- print(bm_umap) -->

<!-- # bm_3classes <- RunUMAP(bm_3classes, dims = 1:28) -->
<!-- #  -->
<!-- # bm_3classes_umap = DimPlot(bm_3classes, reduction = "umap") +  -->
<!-- #     plot_annotation(title = "bm_3classes UMAP") -->
<!-- #  -->
<!-- # print(bm_3classes_umap) -->
<!-- ``` -->


# Annotations with singleR

```{r, include = FALSE}
library(SingleR)
library(celldex)
library(Seurat)
library(here)
library(patchwork)
library(scRNAseq)
library(scuttle)
library(tidyverse)
library(ggsci)

outdir = here("output")

fig.dir = here("figures")
# source(here("scripts", "marker_genes.R"))

# hpca.se <- HumanPrimaryCellAtlasData()

mouse.se = celldex::MouseRNAseqData()

mouse.imm = celldex::ImmGenData()

source(here("scripts", "eos_markers.R"))
```

<!-- ```{r load-combined, echo=FALSE} -->
<!-- # bm.combined = readRDS(file = file.path(outdir, "bm.combined.rds")) -->
<!-- bm = readRDS(file = file.path(outdir, "bm_final.rds")) -->
<!-- ``` -->

## Annotations with MouseRNAseqData

```{r single-R, echo=FALSE}
bm_singleR = SingleR(test = as.SingleCellExperiment(bm), ref = mouse.se, 
                       assay.type.test = 1, labels = mouse.se$label.main)

# bm_singleR_fine = SingleR(test = as.SingleCellExperiment(bm), ref = mouse.se, 
#                        assay.type.test = 1, labels = mouse.se$label.fine)
```

```{r singleR-annot-plots, echo=FALSE}
unrelated_cells = c("Adipocytes", "Cardiomyocytes", "Epithelial cells",
                    "Fibroblasts", "Fibroblasts", "Microglia", "Neurons",
                    "Oligodendrocytes", "Hepatocytes")

bm_labels = bm_singleR$pruned.labels
bm_labels [bm_labels %in% unrelated_cells ] = "Other"
bm_labels = as.factor(setNames(bm_labels, names(bm@active.ident)))
bm$rnaseq_labels = bm_labels

bm@active.ident = bm$rnaseq_labels

all_levels =levels(bm_labels)
all_levels = c("B cells", "Endothelial cells", "Dendritic cells", "Erythrocytes",
               "Monocytes", "Granulocytes", "NK cells", "Macrophages", "Other")

all_colors = setNames(pal_jco()(length(all_levels)), all_levels)

all_colors = setNames(pal_d3("category20")(length(all_colors)), all_levels)

bm_singleR_annotated = DimPlot(bm, reduction = "umap", 
                                 label = FALSE, pt.size = 0.4,
                                 cols = all_colors# , split.by = "sample"
                               ) + 
    plot_annotation(title = "bm - singleR RNA-seq annotated cell types")
print(bm_singleR_annotated)

# ggsave(bm_singleR_annotated, 
#        filename = file.path(fig.dir, "mousernaseq_annotation.umap.png"),
#        width = 8, height = 3.5)
```

## Annotations with MouseRNAseqData ImmGen

```{r}
bm_singleR_immgen = SingleR(test = as.SingleCellExperiment(bm), ref = mouse.imm, 
                       assay.type.test = 1, labels = mouse.imm$label.main)

# bm_singleR_immgen_fine = SingleR(test = as.SingleCellExperiment(bm), ref = mouse.imm, 
#                        assay.type.test = 1, labels = mouse.imm$label.fine)
```

```{r}
imm_gen_cell_counts = bm_singleR_immgen$pruned.labels %>% table() 
cells_few = names(which(imm_gen_cell_counts <10))

bm_immgen_labels = bm_singleR_immgen$pruned.labels
bm_immgen_labels [bm_immgen_labels %in% cells_few ] = "Other"

bm_immgen_labels [bm_immgen_labels == "Tgd" ] = "T cells"
bm_immgen_labels [bm_immgen_labels == "B cells, pro" ] = "B cells"

# bm_immgen_labels = droplevels(bm_immgen_labels)

bm_immgen_labels = as.factor(setNames(bm_immgen_labels, names(bm@active.ident)))
bm$immgen_labels = bm_immgen_labels
bm@active.ident = bm$immgen_labels

all_immgen_levels =levels(bm_immgen_labels)
all_immgen_levels = c("B cells", "Basophils", "DC", "ILC", "Monocytes",
                      "Neutrophils", "NK cells", "NKT", "Other",
                      "T cells", "Stem cells")

all_immgen_colors = setNames(pal_jco()(length(all_immgen_levels)), all_immgen_levels)

all_immgen_colors = setNames(pal_d3("category20")(length(all_immgen_levels)), all_immgen_levels)

bm_singleR_immgen_annotated = DimPlot(bm, reduction = "umap", 
                                 label = FALSE, pt.size = 0.4,
                                 cols = all_immgen_colors# , split.by = "sample"
                                 ) + 
    plot_annotation(title = "bm - singleR ImmGen annotated cell types")
print(bm_singleR_immgen_annotated)
ggsave(bm_singleR_immgen_annotated, 
       filename = file.path(fig.dir, "immgen_annotation.umap.png"),
       width = 8, height = 3.5)
```


```{r fig.width = 10, fig.height = 5}
pp = ggpubr::ggarrange(bm_singleR_annotated, bm_singleR_immgen_annotated)
print(pp)
# ggsave(pp, 
#        filename = file.path(fig.dir, "immgen_rnaseq_annotation.umap.png"),
#        width = 13, height = 5)
```


### Checking eosinophil markers 

#### Eosinophil markers

```{r, fig.width = 10, fig.height = 6}
FeaturePlot(bm, features = eos_markers, pt.size = 0.3)
```


#### Eosinophil progenitor markers

```{r, fig.width = 10, fig.height = 6}
FeaturePlot(bm, features = eos_progenitors, ncol = 3, pt.size = 0.3)
```

```{r saving-data, include = FALSE}
saveRDS(bm, file = file.path(outdir, "bm_pilot.rds") )

# saveRDS(bm_3classes, file = file.path(outdir, "bm_3classes.rds") )

bm.split = SplitObject(bm, split.by = "sample")
saveRDS(bm.split, file = file.path(outdir, "bm_split.rds"))
```

```{r}
bm@active.assay = "RNA"
# SaveH5Seurat(bm, file = file.path(outdir, "bm_pilot_singlet.h5Seurat"), )
# Convert(file.path(outdir, "bm_pilot_singlet.h5Seurat"), dest = "h5ad")
sceasy::convertFormat(bm, from="seurat", to="anndata",
                      outFile=file.path(outdir, "bm_pilot_singlet.h5ad"))
```

